Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-04-29 concurrency/CONCUR 2023/paper.tex
finished proof of give-up traverse abd added in appendix

%maybe inline this instead of making it a figure?
\begin{figure}[ht]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
			\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
			\begin{array}{c} \texttt{pn} \mapsto (\texttt{p, n}) \ \ast 
				\ (\mathit{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL}))  \ \ast \ \nodeboxrep(\texttt{pn->n'}) \\  
				\ast \ \mathsf{R}(\texttt{pn->n'})\ \ast \ (\texttt{k} \in \mathsf{range}(\texttt{n'}))
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the lock-coupling template}
	\label{fig:traverse_lock}
\end{figure}

@@

In the specification, the local precondition includes both the node handle $\nodeboxrep(\texttt{n})$ and the contents $\mathsf{R}(\texttt{n})$. The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}. The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a boolean variable $\mathit{res}$ indicating whether \texttt{n'->t} is \texttt{NULL} or not. The variable $\mathit{res}$ is $\mathsf{true}$ if $\texttt{flag}$ returns 1, and $\mathsf{false}$ if 0, mirroring its boolean state. The key to the correctness of the \texttt{traverse} function the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range}(\texttt{n'})\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \  \mathsf{R}(\texttt{n'}) \ \ast \ \treerep\ m
\end{mathpar} %\wm{Where does \lstinline{range} come from here?}

@@

				$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} = \texttt{NULL}  \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$

@@

@@

						$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \texttt{n'} = \texttt{n} \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots    \end{array}\right\}$

@@

@@

				(\mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast (\texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots)

@@

	\begin{array}{c} \texttt{pn} \mapsto (\texttt{p, n'}) \ \ast 
		\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \nodeboxrep(\texttt{n'}) \ \ast \
		\\ \mathsf{R}(\texttt{n'})\ \ast \ (\texttt{k} \in \mathsf{range}(\texttt{n'}))

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle
			\begin{array}{c}
				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \infp (\texttt{n}) \ \big| \ \treerep\ m
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \ \texttt{n'}.
			\begin{array}{c}
				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL}))  \ \ast \ 
				\\ \infp(\texttt{pn->n'}) \ \ast \ \mathsf{R}(\texttt{pn->n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}

The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup},  is simpler than that of the lock-coupling template. The local precondition include any locks, and consists only of $\infp(\texttt{pn->n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$. As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \lstinline{n'}'s lock and returning its contents $\mathsf{R}(\texttt{pn->n'})$. Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.

@@

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
\end{mathpar}
With this invariant in hand, the proof of \lstinline{traverse} proceeds as follows. We start with the $\infp (\texttt{pn->n})$ predicate from the local precondition. After acquiring the lock of the node $\texttt{pn->n}$ (line 5 in Figure \ref{traverse_giveup_b}), we obtain the resource $\mathsf{R}(\texttt{pn->n})$, which allows us to perform the \lstinline{inRange} check on line 7. If the check fails (lines 20-23), we immediately release the lock and start over from the beginning. Otherwise, we proceed as before: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \lstinline{k} belongs (and can satisfy the postcondition); if \lstinline{findNext} returns 0, we have found the node containing \lstinline{k} (and can satisfy the postcondition); otherwise, we release our lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \lstinline{findNext}. In this case, we do not hold the lock on the new node, but only the $\infp$ assertion indicating that the new node is in the data structure. % and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}. Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (line 8 in Figure \ref{traverse_giveup_b}). At this point, we can prove the post-condition using the specification, which indicates that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$.

@@

\newpage
\bibliography{../sources}

%\appendix

%\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}

>>>>

%maybe inline this instead of making it a figure?
\begin{figure}[ht]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
			\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
			\begin{array}{c} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
				\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the lock-coupling template}
	\label{fig:traverse_lock}
\end{figure}

@@

In the specification, the local precondition includes both the node handle $\nodeboxrep(\texttt{n})$ and the contents $\mathsf{R}(\texttt{n})$. The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}. The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a boolean variable $\mathit{res}$ indicating whether \texttt{n'->t} is \texttt{NULL} or not. The variable $\mathit{res}$ is $\mathsf{true}$ if $\texttt{flag}$ returns 1, and $\mathsf{false}$ if 0, mirroring its boolean state. The key to the correctness of the \texttt{traverse} function the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range}(\texttt{n'})\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \  \mathsf{R}(\texttt{n'}) \ \ast \ \treerep\ m
\end{mathpar} %\wm{Where does \lstinline{range} come from here?}

@@

@@

						$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} = \texttt{NULL}  \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$

@@

@@

						$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \texttt{n'} = \texttt{n} \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots    \end{array}\right\}$

@@

		(\mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots)

@@

	\begin{array}{c} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
		\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle
			\begin{array}{c}
				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \infp (\texttt{n}) \ \big| \ \treerep\ m
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \ \texttt{n'}.
			\begin{array}{c}
				\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
				\\ \infp(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}

The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup},  is simpler than that of the lock-coupling template. The local precondition include any locks, and consists only of $\infp(\texttt{n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$. As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \lstinline{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$. Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.

@@

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ \texttt{p q n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \ \infp (\texttt{q}) \ast \infp (\texttt{n'})  \ \ast \ \treerep\ m
\end{mathpar}
With this invariant in hand, the proof of \lstinline{traverse} proceeds as follows. We start with the $\infp (\texttt{n})$ predicate from the local precondition. After acquiring the lock of the node $\texttt{n}$ (line 5 in Figure \ref{proof_giveup_traverse} in Appendix \ref{sec:apd_giveup_proof}), we obtain the resource $\mathsf{R}(\texttt{n})$, which allows us to perform the \lstinline{inRange} check on line 7. If the check fails (lines 20-23), we immediately release the lock and start over from the beginning. Otherwise, we proceed as before: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \lstinline{k} belongs (and can satisfy the postcondition); if \lstinline{findNext} returns 0, we have found the node containing \lstinline{k} (and can satisfy the postcondition); otherwise, we release our lock and re-establish $\mathsf{traverse\_inv}$ for the new node indicated by \lstinline{findNext}. In this case, we do not hold the lock on the new node, but only the $\infp$ assertion indicating that the new node is in the data structure. % and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}. Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (line 8 in Figure \ref{traverse_giveup_b}). At this point, we can prove the post-condition using the specification, which indicates that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$.

@@

\bibliography{../sources}

\appendix

\section{Give-up Proof}
\label{sec:apd_giveup_proof}
	\forall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \infp (\texttt{n}) \ \big| \ \treerep\ m
			node_t *r = (pn->n);
				\infp (\texttt{n})  \ast \infp (\texttt{r}) \ast \treerep\ m 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
				$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots \end{array}\right\}$
				pn->p = pn->n; $\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots \end{array}\right\}$
				if (inRange(pn, k) == 1){
					$\color{blue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast  \cdots \end{array}\right\}$
					$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} = \texttt{NULL}  \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$
							(\mathit{res'} = \mathsf{false} \ast \texttt{n'} = \texttt{n} \ast  \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots )\  \lor \\ 
							(\mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \infp (\texttt{n'}) \ast \cdots)   \end{array}\right\}$
							$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL}  \ast \texttt{n'} = \texttt{n} \ast  \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots   \end{array}\right\}$
						else
						$\color{blue} \left\{\begin{array}{l} \mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n->t} \neq \texttt{NULL} \ast \infp (\texttt{n'}) \ast \cdots    \end{array}\right\}$
						$\color{blue} \left\{\begin{array}{l} \mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n->t} \neq \texttt{NULL} \ast \infp (\texttt{n'}) \ast \cdots    \end{array}\right\}$
					$\color{blue} \left\{\begin{array}{l} \texttt{k} \not\in \mathsf{range}(\texttt{n}) \ast \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast  \cdots \end{array}\right\}$
					$\color{blue} \left\{\begin{array}{l} \texttt{k} \not\in \mathsf{range}(\texttt{n}) \ast \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast  \cdots \end{array}\right\}$
					pn->n = r;
					$\color{blue} \left\{\begin{array}{l} \texttt{k} \not\in \mathsf{range}(\texttt{n}) \ast  \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{r, r}) \ast \infp (\texttt{r}) \ast \cdots    \end{array}\right\}$
			$\color{blue} \left\{\begin{array}{l} (\mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots) \  \lor \\
				(\mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots)
	\begin{array}{c}
		\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
		\\ \infp(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
	\caption{The \texttt{traverse} function in give-up template annotated with separation logic specification}
	\label{proof_giveup_traverse}
